A comps-based approach for interpreting tree-based predictions

### 2025-05-19 articles/arxiv/paper.tex
fine-tune data section

There were two sources for our data. The NFL data for ESPN's Total Quarterback Rating (QBR) came from \href{https://pro-football-reference.com}{pro-football-reference.com}. For the corresponding college statistics, we used data from \href{https://sports-reference.com}{sports-reference.com}. The college statistics only include schools that are a part of the NCAA Division 1 Football Bowl Subdivision (FBS). We were able to obtain college statistics dating back to 1984. However, we were constrained by the ESPN QBR metric, which has calculations starting at the 2006 season.

The data from Sports Reference contained player passing statistics for each season. It contained categories for player, team, conference, games, completions, attempts, yards, touchdowns, touchdown percentage, interceptions, interception percentage, yards per attempt, adjusted yards per attempt, yards per completion, yards per game, passing efficiency rating, awards, and season. In addition to the passing statistics, we included player rushing statistics for each season from Sports Reference.

We obtained NFL season passing statistics from Pro Football Reference for the 2006-2024 seasons. The resulting data frame contained columns for 28 different passing metrics. However, we only used QBR and overall pick from this model.

\subsection{ESPN's Total Quarterback Rating}

The NFL response variable we used in this paper is ESPN's Total Quarterback Rating (QBR). It is an EPA-based metric that is adjusted for home-field advantage, defensive strength, and ``garbage time'' and then scaled from 0-100 \citep{noauthor_how_2016}. QBR provides a more holistic evaluation of quarterbacks in the NFL while also considering other factors affecting performance. Additionally, a 0-100 scale is important in this context because it offers an interpretable standard frame of reference. It also allows us to assign a meaningful value to quarterbacks who never played in the NFL, compared to raw EPA (where the lack of such scale results in negative values for some players). For example, it would not make much sense for a quarterback who receives playing time in the NFL to have a lower value than a quarterback who never made it to the NFL, but this would happen if using raw EPA as the response variable. Using QBR as the response variable addresses this issue while incorporating other important aspects of game-time performance.

>>>>

All data used for this study are publicly available from Sports Reference \citep{sports_reference_sports_2025}. Our dataset includes 2,114 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who averaged more than six games played per season and more than five passing attempts per season. For each of these quarterbacks, we obtained the following data from their college career: games played per season, passing attempts per season, passing yards per season, passing yards per attempt, passing touchdowns per season, interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season conference, final-season games played, final-season passing yards per attempt, and final-season passer rating.

In addition to college statistics, we obtained each player's passing attempts and Total Quarterback Rating (QBR) \citep{burke_how_2016} from each season between 2006 and 2024 (inclusive). Only available since 2006, QBR is based on expected points added (EPA), which is the change in drive scoring expectation before and after each play. It considers each quarterback's share of their team's EPA and accounts for home-field advantage, defensive strength, and garbage time. The range of QBR is from 0 to 100, and \citet{burke_how_2016} suggests thinking of it as a percentile of performance at the individual game level. For each quarterback, we calculate their average QBR per season, and we define a player who has attempted fewer than 10 career NFL passes to have a QBR of zero. Here the scale of QBR is helpful because players who reach the NFL cannot be rated below players who never reach the NFL.

Average QBR per season is the response variable describing the success of quarterback prospects. Because of the date ranges of the data, we do not calculate this perfectly for all of the quarterbacks in our dataset. On one end of the extremes, quarterbacks who finished their college careers in 2001 may have appeared in the NFL but only before 2006. These quarterbacks have a QBR of zero in our dataset. On the other end of the extremes, quarterbacks who finished their college careers in 2023 and have not yet appeared in the NFL may yet do so. These quarterbacks, too, have a QBR of zero in our dataset. This is a limitation of the data.
